Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-02-07 paper.tex

\subsection{Tile State Implementation}
Since workgroups must observe updates from one another, but WebGPU lacks a \emph{volatile}-like qualifier, we rely on atomics to ensure visibility and consistency across workgroups. Because a tile may be in three possible states, two bits are required to record its state alongside the bits needed to store the tile's reduction value. Previous \emph{Chained-Scan} architectures relied on 64-bit atomics and/or memory fences to ensure a consistency between state and value, but WebGPU’s lack of a memory model \john{I'm not sure I would say ``lack of a memory model''. It's not the (lack of) memory model that's the problem. It's specifically that we don't have the operations that give us the behavior that we require. If we had the right operations that did the right thing but didn't have a memory model, we'd be fine. We want to specifically call out what is missing and what that means, and do so as generically as possible (not ``we are missing this specific atomic call'' but instead ``we are missing some sort of atomic call that has the following behavior''). Above, we want to not only say that we don't have a volatile qualifier but also both what that means and why it would be specifically useful if we had it. Doing the latter two things gives valuable ammunition to WebGPU and GPU architects that want to improve the API and hardware.} and 64-bit atomic support makes both infeasible. Instead, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state. \john{I would try to add a sentence or two on why we can't do the obvious thing---having the state in one word and the reduction value in a different word---and why the split value avoids that problem. This is not obvious to the reader.}
\begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{graphics/split.pdf}
\end{figure}

>>>>

\subsection{Tile State Implementation}
In all variants of \emph{Chained-Scan}, there is communication and synchronization between workgroups. The fundamental primitive is \emph{message passing,} meaning that one workgroup posts the result of a communication, then other workgroups read it. This primitive is well studied in the context of memory models. \raph{The earliest publication of the MP litmus test appears to be section 2.1 of "Intel 64 Architecture Memory Ordering White Paper". I'm not sure that's the best cite; the MC mutants paper cites https://arxiv.org/pdf/1207.7264} The key requirement is both control synchronization, meaning the reader does not proceed until the writer sends the message; and data synchronization, meaning that the reader reads the correct value. The latter is especially challenging in relaxed memory models as are employed on GPUs.

Combined control and data synchronization can be accomplished by packing both the tile state and the reduced value into a single word, using relaxed atomic store to send the message, and relaxed atomic load to receive the message, spinning as necessary until the tile state is updated. The semantics of atomics guarantee, even with relaxed memory ordering, that the data is fresh. However, if the tile state and data are stored in two different memory locations, then relaxed atomics are not sufficient and some other sychronization is necessary. On CPUs, it is common to upgrade the memory semantics from relaxed to acquire/release, which then provides the necessary guarantee. \raph{Here's a good place to cite the MP litmus test because that's exactly what it's about.} Unfortunately, on GPUs, acquire/release semantics are neither portable (they are only available on Vulkan, and optional until version 1.3), nor are they necessarily efficient even when they are available; often the implementation involves expensive cache flushes. A device-scoped memory barrier has similar problems. While it would provide the necessary guarantee, it is also not available in WebGPU (or in versions of Metal prior to 3.2), and is expensive.

Without some other mechanism, the requirement to pack tile state and data into a single word limits the data to 30 bits (assuming 2 bits for the tile state). That, in turn, makes it impossible to represent common ordinary 32-bit floating point values without loss of precision. While some GPUs support 64 bit atomics (and it is likely that WebGPU will include them as an optional extension), that is also sadly not portable, and merely raises the payload limit to 62 bits without eliminating it altogether.

A key innovation of this paper is to \emph{split} the reduced values into multiple atomics, each carrying its own copy of the control flags. The message is received when relaxed atomic loads of all the parts agree on their control flag. This way, the payload is reconstructed intact, even in the absence of any guarantees from the memory model stronger than relaxed atomics.

In particular, we use two or four 32-bit values, splitting the reduction value into 16-bit segments while reserving the two most significant bits of each 32-bit value to encode the tile state.
\begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{graphics/split.pdf}
\end{figure}
